% =====================================================================================
% ABSTRACT
% =====================================================================================

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Process mining has emerged as a crucial discipline for understanding and optimizing business processes through event log analysis. However, traditional visualization methods often fail to effectively represent the temporal characteristics of process events, particularly when dealing with highly skewed time distributions commonly found in real-world process data. This thesis presents an innovative approach to process mining visualization through the development of an interactive dashboard that employs violin plots for temporal event analysis.

The core contribution of this work is the systematic solution to the "case-start event problem" in process mining, where events occurring at the beginning of process instances (time = 0) dominate visualizations and obscure meaningful temporal patterns. Our approach introduces intelligent event filtering combined with multiple time transformation methods to reveal hidden process insights across diverse domains.

We developed a comprehensive multi-dataset dashboard that supports real-time analysis across four distinct process domains: government (Traffic Fine Management with 561K events), finance (BPI Challenge 2012 and 2017 with up to 1.2M events), and healthcare (Sepsis Cases with 15K events). The system provides seven different time transformation methods, dynamic event sorting capabilities, and configurable analysis depth to accommodate various research needs.

Our evaluation demonstrates the effectiveness of violin plots in revealing temporal patterns that remain hidden in traditional visualizations. The dashboard successfully handles datasets ranging from 15K to 1.2M events with consistent performance, enabling researchers to conduct cross-domain process analysis within a single interface. Key findings include the identification of bi-modal payment patterns in government processes, process bottlenecks in financial workflows, and time-critical decision patterns in healthcare.

The research contributions include: (1) a methodological framework for intelligent event filtering in temporal process analysis, (2) a modular multi-dataset architecture supporting real-time cross-domain comparison, (3) empirical validation across four diverse process types demonstrating broad applicability, and (4) an open-source dashboard making advanced process mining visualization accessible to the research community.

Results show that our approach reveals process insights not visible through conventional methods, with implications for process optimization in healthcare, finance, and government domains. The dashboard achieves sub-second response times for all transformations and supports seamless dataset switching, making it suitable for both research and practical applications.

\textbf{Keywords:} Process Mining, Data Visualization, Violin Plots, Interactive Dashboard, Temporal Analysis, Event Logs, Cross-Domain Analysis

\cleardoublepage
